---
title: "Final Project - Intro to Data Science - 6101"
author: "We-R-4(3)"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r include=FALSE}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
library(readr)
library(dplyr)
library(ggplot2)
library(car)
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r}
credit1 <- read_csv('application_record.csv')
credit2 <- read_csv('credit_record.csv')
```

# Preparation and Initial Analysis

The data set originally consists of 2 separate csv files that are connected to each other via the 'ID' column. The first csv file, application record.csv, consists of personal and socio-economic data that is related to an applicant and can help a bank decide whether to issue them a credit card or not. The second csv file, credit_record.csv, consists of past credit card usage behavior of an applicant. It consists of information such as defaults, payment due dates etc. We will perform a 'left join' to merge these two files and so that we are working with the right applicants and will get rid of the extra rows depending on the Null values. 

```{r}
df <- merge(credit1, credit2, by = 'ID', all.x = TRUE)
```

Checking NA values in the updated dataframe and performing cleanup in the dataframe columns. 

```{r}
sum(is.na(df$OCCUPATION_TYPE)) # 'Occupation_type' has nearly 350000 na values in the original dataframe, along with 'Status' and 'months_balance'.
# Removing the NA values from the above columns:
new_df <- na.omit(df, c("STATUS", "MONTHS_BALANCE", "OCCUPATION_TYPE"))
# Making sure the dataframe has distinct values: 
df2 <- distinct(new_df, ID, .keep_all = TRUE)

```
The above generated data frame(df2) has 25134 rows and 20 variables. 

```{r}


```



```{r}


```



```{r}


```



```{r}


```



```{r}


```