---
title: "Final Project - Intro to Data Science - 6101"
author: "We-R-4(3)"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: true
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r include=FALSE}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
library(readr)
library(dplyr)
library(ggplot2)
library(car)
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r}
credit1 <- read_csv('application_record.csv')
credit2 <- read_csv('credit_record.csv')
```

# Preparation, Initial Analysis, and Data Tidying:

The data set originally consists of 2 separate csv files that are connected to each other via the 'ID' column. The first csv file, application record.csv, consists of personal and socio-economic data that is related to an applicant and can help a bank decide whether to issue them a credit card or not. The second csv file, credit_record.csv, consists of past credit card usage behavior of an applicant. It consists of information such as defaults, payment due dates etc. We will perform a 'left join' to merge these two files and so that we are working with the right applicants and will get rid of the extra rows depending on the Null values. 

```{r}
df <- merge(credit1, credit2, by = 'ID', all.x = TRUE)
```

Checking NA values in the updated dataframe and performing cleanup in the dataframe columns. 

```{r}
sum(is.na(df$OCCUPATION_TYPE)) # 'Occupation_type' has nearly 350000 na values in the original dataframe, along with 'Status' and 'months_balance'.
# Removing the NA values from the above columns:
new_df <- na.omit(df, c("STATUS", "MONTHS_BALANCE", "OCCUPATION_TYPE"))
# Making sure the dataframe has distinct values: 
df2 <- distinct(new_df, ID, .keep_all = TRUE)

```
The above generated data frame(df2) has 25134 rows and 20 variables. 

Removing excess columns or ones that do not pertain to the target variable. 

```{r}
df2 <- subset( df2, select = -c(FLAG_EMAIL, FLAG_PHONE, FLAG_WORK_PHONE, DAYS_EMPLOYED, DAYS_BIRTH, ID, OCCUPATION_TYPE))
df2$ID <- 1:nrow(df2)
df2 <- df2 %>% relocate(ID, .before = CODE_GENDER)
# Renaming columns to more generic values:
df2 <- df2 %>% rename(gender = CODE_GENDER, owns_car = FLAG_OWN_CAR, owns_realty = FLAG_OWN_REALTY, children = CNT_CHILDREN, total_income = AMT_INCOME_TOTAL, employment_type = NAME_INCOME_TYPE, education_status = NAME_EDUCATION_TYPE, marital_status = NAME_FAMILY_STATUS, housing_type = NAME_HOUSING_TYPE, owns_mobile = FLAG_MOBIL, family_members = CNT_FAM_MEMBERS, months_due = MONTHS_BALANCE, status = STATUS)

df2$months_due <- df2$months_due * (-1)

```

Converting all categorical values to ordinal/numeric values for further computation:


```{r}
df2$gender <- replace(df2$gender, df2$gender=="M", 1)
df2$gender <- replace(df2$gender, df2$gender=="F", 0)
df2$gender <- as.numeric(df2$gender)

df2$owns_car <- replace(df2$owns_car, df2$owns_car=="Y", 1)
df2$owns_car <- replace(df2$owns_car, df2$owns_car=="N", 0)
df2$owns_car <- as.numeric(df2$owns_car)

df2$owns_realty <- replace(df2$owns_realty, df2$owns_realty=="Y", 1)
df2$owns_realty <- replace(df2$owns_realty, df2$owns_realty=="N", 0)
df2$owns_realty <- as.numeric(df2$owns_realty)

df2$employment_type <- replace(df2$employment_type, df2$employment_type=="Working", 0)
df2$employment_type <- replace(df2$employment_type, df2$employment_type=="Commercial associate", 1)
df2$employment_type <- replace(df2$employment_type, df2$employment_type=="State servant", 2)
df2$employment_type <- replace(df2$employment_type, df2$employment_type=="Student", 3)
df2$employment_type <- replace(df2$employment_type, df2$employment_type=="Pensioner", 4)
df2$employment_type <- as.numeric(df2$employment_type)

df2$education_status <- replace(df2$education_status, df2$education_status=="Secondary / secondary special", 0)
df2$education_status <- replace(df2$education_status, df2$education_status=="Higher education", 1)
df2$education_status <- replace(df2$education_status, df2$education_status=="Incomplete higher", 2)
df2$education_status <- replace(df2$education_status, df2$education_status=="Lower secondary", 3)
df2$education_status <- replace(df2$education_status, df2$education_status=="Academic degree", 4)
df2$education_status <- as.numeric(df2$education_status)

df2$marital_status <- replace(df2$marital_status, df2$marital_status=="Married", 0)
df2$marital_status <- replace(df2$marital_status, df2$marital_status=="Single / not married", 1)
df2$marital_status <- replace(df2$marital_status, df2$marital_status=="Civil marriage", 2)
df2$marital_status <- replace(df2$marital_status, df2$marital_status=="Separated", 3)
df2$marital_status <- replace(df2$marital_status, df2$marital_status=="Widow", 4)
df2$marital_status <- as.numeric(df2$marital_status)

df2$housing_type <- replace(df2$housing_type, df2$housing_type=="House / apartment", 0)
df2$housing_type <- replace(df2$housing_type, df2$housing_type=="Rented apartment", 1)
df2$housing_type <- replace(df2$housing_type, df2$housing_type=="Municipal apartment", 2)
df2$housing_type <- replace(df2$housing_type, df2$housing_type=="With parents", 3)
df2$housing_type <- replace(df2$housing_type, df2$housing_type=="Co-op apartment", 4)
df2$housing_type <- replace(df2$housing_type, df2$housing_type=="Office apartment", 5)
df2$housing_type <- as.numeric(df2$housing_type)

df2$status <- replace(df2$status, df2$status=="0", 0)
df2$status <- replace(df2$status, df2$status=="X", 0)
df2$status <- replace(df2$status, df2$status=="C", 0)
df2$status <- replace(df2$status, df2$status=="1", 0)
df2$status <- replace(df2$status, df2$status=="2", 1)
df2$status <- replace(df2$status, df2$status=="3", 1)
df2$status <- replace(df2$status, df2$status=="4", 1)
df2$status <- replace(df2$status, df2$status=="5", 1)
df2$status <- as.numeric(df2$status)

```

The target variable 'status' has the following values:
0: 1-29 days past due
1: 30-59 days past due
2: 60-89 days overdue
3: 90-119 days overdue
4: 120-149 days overdue
5: Overdue or bad debts, write-offs for more than 150 days 
C: paid off that month 
X: No loan for the month

Any Customers that are overdue for more than 60 days are marked as `risk` or 1 in the target variable. Other customers are marked as 0. 
The target variable status has a binary outcome, 1 or 0, predicting if an applicant will be a `risk` or not. 

```{r}


```